---
title: "Figure_2_stability_groups"
output: html_document
date: '2025-04-13'
editor_options:
  chunk_output_type: console
---

# Figure 2 stability groups
#
# This script regenerates Fig. 2C, E, G and Supplementary Fig. 2B.
#
# Expected repository structure:
# Figure_2/
#   ├── data/
#   │   └── merged_data_GC_filtered.csv
#   ├── results/
#   └── scripts/
#       └── Figure_2_stability_groups.Rmd
#
# This script assumes it is run from the scripts/ folder.
#
# Input file:
# merged_data_GC_filtered.csv is a filtered table in which introns were
# retained only if they had:
# - at least 5 non-missing nuclear PIR values across the time course
# - Nuc_UT_TPM_mean >= 5

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(pheatmap)
  library(RColorBrewer)
  library(ggplot2)
  library(tidyr)
  library(patchwork)
  library(readr)
  library(purrr)
  library(broom)
})
```

# Input and output paths
```{r}
input_file <- "../data/merged_data_GC_filtered.csv"
output_dir <- "../results"

dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)
```

# Load data
```{r}
merged_data_GC_filtered <- read.csv(input_file, stringsAsFactors = FALSE)

# Set working directory for all output files written below.
setwd(output_dir)

merged_data_GC_filtered <- merged_data_GC_filtered %>% 
  mutate(
    Nuc_UT_mean = as.numeric(as.character(Nuc_UT_mean)),
    Nuc_30_mean = as.numeric(as.character(Nuc_30_mean)),
    Nuc_2h_mean = as.numeric(as.character(Nuc_2h_mean)),
    Nuc_4h_mean = as.numeric(as.character(Nuc_4h_mean)),
    Cyt_UT_mean = as.numeric(as.character(Cyt_UT_mean)),
    Cyt_30_mean = as.numeric(as.character(Cyt_30_mean)),
    Cyt_2h_mean = as.numeric(as.character(Cyt_2h_mean)),
    Cyt_4h_mean = as.numeric(as.character(Cyt_4h_mean)))

merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(
    Nuc_A_UT_IR_stability = Nuc_A_UT_1.x / Nuc_UT_mean,
    Nuc_B_UT_IR_stability = Nuc_B_UT_1.x / Nuc_UT_mean,
    
    Nuc_A_30_IR_stability = Nuc_A_30_1.x / Nuc_UT_mean,
    Nuc_B_30_IR_stability = Nuc_B_30_1.x / Nuc_UT_mean,
    
    Nuc_A_2h_IR_stability = Nuc_A_2h_1.x / Nuc_UT_mean,
    Nuc_B_2h_IR_stability = Nuc_B_2h_1.x / Nuc_UT_mean,
    
    Nuc_A_4h_IR_stability = Nuc_A_4h_1.x / Nuc_UT_mean,
    Nuc_B_4h_IR_stability = Nuc_B_4h_1.x / Nuc_UT_mean
  )
```

```{r}
# Create new columns comparing the stability of RNA in the nucleus
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Nuc_UT_TPM_stability = Nuc_UT_TPM_mean / Nuc_UT_TPM_mean,  # Reference of 1 for UT
         Nuc_30_TPM_stability = Nuc_30_TPM_mean / Nuc_UT_TPM_mean,
         Nuc_2h_TPM_stability = Nuc_2h_TPM_mean / Nuc_UT_TPM_mean,
         Nuc_4h_TPM_stability = Nuc_4h_TPM_mean / Nuc_UT_TPM_mean)

# Create new columns comparing the stability of RNA in the cytoplasm
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_mean = as.numeric(as.character(Cyt_UT_mean)),
         Cyt_30_mean = as.numeric(as.character(Cyt_30_mean)),
         Cyt_2h_mean = as.numeric(as.character(Cyt_2h_mean)),
         Cyt_4h_mean = as.numeric(as.character(Cyt_4h_mean)))

merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_TPM_stability = Cyt_UT_TPM_mean / Cyt_UT_TPM_mean,  # Reference of 1 for UT
         Cyt_30_TPM_stability = Cyt_30_TPM_mean / Cyt_UT_TPM_mean,
         Cyt_2h_TPM_stability = Cyt_2h_TPM_mean / Cyt_UT_TPM_mean,
         Cyt_4h_TPM_stability = Cyt_4h_TPM_mean / Cyt_UT_TPM_mean)

# Create new columns comparing the stability of IR in the nucleus
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Nuc_UT_IR_stability = Nuc_UT_mean / Nuc_UT_mean,  # Reference of 1 for UT
         Nuc_30_IR_stability = Nuc_30_mean / Nuc_UT_mean,
         Nuc_2h_IR_stability = Nuc_2h_mean / Nuc_UT_mean,
         Nuc_4h_IR_stability = Nuc_4h_mean / Nuc_UT_mean)

# Create new columns comparing the stability of IR in the cytoplasm
merged_data_GC_filtered <- merged_data_GC_filtered %>%
  mutate(Cyt_UT_IR_stability = Cyt_UT_mean / Cyt_UT_mean,  # Reference of 1 for UT
         Cyt_30_IR_stability = Cyt_30_mean / Cyt_UT_mean,
         Cyt_2h_IR_stability = Cyt_2h_mean / Cyt_UT_mean,
         Cyt_4h_IR_stability = Cyt_4h_mean / Cyt_UT_mean)
```

# For gene-orientated studies sometimes you need to filter the data to keep only the intron with the highest Nuc_UT_mean for each gene
```{r}
merged_data_GC_filtered_GENE <- merged_data_GC_filtered %>%
  group_by(GENE) %>%
  top_n(1, Nuc_UT_mean) %>%  # Select the top 1 row with the highest Nuc_UT_mean per GENE
  ungroup()
```

# Assign the stability to introns
```{r}
#--------------------------------------------#
#### Assign stability to introns
#--------------------------------------------#
IR_UT <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 & 
      (
        (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
        (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
        (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
      )
  ) %>%
  mutate(stability = "IR")

stableR4h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_4h_IR_stability >= 0.5 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability >= 0.5 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    ) &
    Nuc_30_mean >= 25 &
    Nuc_2h_mean >= 25 &
    Nuc_4h_mean >= 25
  ) %>% mutate(stability = "stable 4h")

unstableR30min <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_30_IR_stability < 0.5 &
    Nuc_2h_IR_stability < 0.5 &
    Nuc_4h_IR_stability < 0.5 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    )) %>%
  mutate(stability = "unstable30min")

unstableR2h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability < 0.5 &
    Nuc_4h_IR_stability < 0.5 &
    Nuc_30_mean >= 25 &
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x)))
  ) %>% mutate(stability = "unstable2h")

unstableR4h <- merged_data_GC_filtered %>%
  filter(
    Nuc_UT_mean >= 30 &
    Nuc_4h_IR_stability < 0.5 &
    Nuc_30_IR_stability >= 0.5 &
    Nuc_2h_IR_stability >= 0.5 &
    Nuc_30_mean >= 25 &  
    Nuc_2h_mean >= 25 & 
    (
      (Nuc_A_UT_1.x >= 25 & Nuc_B_UT_1.x >= 25) |
      (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 25) |
      (Nuc_A_UT_1.x >= 25 & is.na(Nuc_B_UT_1.x))
    )
  ) %>% mutate(stability = "unstable 4h")

noIR <- merged_data_GC_filtered %>%
  filter(Nuc_UT_mean < 30) %>%
  mutate(stability = "no IR")

##no IR GENE
noIR_GENE <- merged_data_GC_filtered_GENE %>%
  filter((Nuc_UT_mean) < 30) %>%
  mutate(stability = "no_IR_GENE")

noIR_GENE <- noIR_GENE %>%
  distinct(GENE, .keep_all = TRUE)
sum(is.na(noIR_GENE$Nuc_UT_mean))

combined_data <- bind_rows(stableR4h, unstableR4h, unstableR30min, unstableR2h, noIR)
```

### same settings at 50%
```{r}
# IR at UT (50 cutoff)
IR_UT_50cut <- subset(merged_data_GC_filtered,
                      Nuc_UT_mean >= 50 &
                      (
                        (Nuc_A_UT_1.x >= 50 & Nuc_B_UT_1.x >= 50) |
                        (is.na(Nuc_A_UT_1.x) & Nuc_B_UT_1.x >= 50) |
                        (Nuc_A_UT_1.x >= 50 & is.na(Nuc_B_UT_1.x))
                      )) %>% mutate(stability = "IR")

# Stable at 4h (50 cutoff)
stableR4h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_4h_IR_stability >= 0.5 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability >= 0.5 
  ) %>%
  mutate(stability = "stable 4h")

# Unstable at 30 min (50 cutoff)
unstableR30min_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_30_IR_stability < 0.5 &
  Nuc_2h_IR_stability < 0.5 &
  Nuc_4h_IR_stability < 0.5
  ) %>%
  mutate(stability = "unstable30min")

# Unstable at 2h (50 cutoff)
unstableR2h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability < 0.5 &
  Nuc_4h_IR_stability < 0.5 #&
  ) %>%
  mutate(stability = "unstable2h") 

# Unstable at 4h (50 cutoff)
unstableR4h_50cut <- subset(
  merged_data_GC_filtered,
  Nuc_UT_mean >= 50 &
  Nuc_4h_IR_stability < 0.5 &
  Nuc_30_IR_stability >= 0.5 &
  Nuc_2h_IR_stability >= 0.5 #&
) %>% mutate(stability = "unstable 4h") 

noIR_50cut <- merged_data_GC_filtered %>%
  filter(Nuc_UT_mean < 50) %>%
  mutate(stability = "no IR")

# Final combined dataset for 50% IR cutoff
combined_data_50cut <- bind_rows(
  stableR4h_50cut,
  unstableR4h_50cut,
  unstableR30min_50cut,
  unstableR2h_50cut,
  noIR_50cut
)
```

### Intron differences with different cutoff criteria 30%:
```{r}
combined_dataTPMno100 <- combined_data %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | #keep is.na because otherwise they get moved
  (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

summary_table_no100 <- combined_dataTPMno100 %>%
  group_by(stability) %>%  # Group the data by the stability group
  summarise(
    count         = n(),  # Number of observations per group
    mean_GC       = mean(GC_Content, na.rm = TRUE),
    median_GC     = median(GC_Content, na.rm = TRUE),
    sd_GC         = sd(GC_Content, na.rm = TRUE),
    min_GC        = min(GC_Content, na.rm = TRUE),
    max_GC        = max(GC_Content, na.rm = TRUE),
    mean_LENGTH   = mean(LENGTH, na.rm = TRUE),
    median_LENGTH = median(LENGTH, na.rm = TRUE),
    sd_LENGTH     = sd(LENGTH, na.rm = TRUE),
    min_LENGTH    = min(LENGTH, na.rm = TRUE),
    max_LENGTH    = max(LENGTH, na.rm = TRUE))

print(summary_table_no100)
```

### same summary tables for 50%
```{r}
# Create TPM-filtered datasets from 50% IR cutoff classifications
combined_dataTPMno100_50cut <- combined_data_50cut %>%
   filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | 
         (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

summary_table_no100_50cut <- combined_dataTPMno100_50cut %>%
  group_by(stability) %>%
  summarise(
    count         = n(),
    mean_GC       = mean(GC_Content, na.rm = TRUE),
    median_GC     = median(GC_Content, na.rm = TRUE),
    sd_GC         = sd(GC_Content, na.rm = TRUE),
    min_GC        = min(GC_Content, na.rm = TRUE),
    max_GC        = max(GC_Content, na.rm = TRUE),
    mean_LENGTH   = mean(LENGTH, na.rm = TRUE),
    median_LENGTH = median(LENGTH, na.rm = TRUE),
    sd_LENGTH     = sd(LENGTH, na.rm = TRUE),
    min_LENGTH    = min(LENGTH, na.rm = TRUE),
    max_LENGTH    = max(LENGTH, na.rm = TRUE)
  )

print(summary_table_no100_50cut)
```

### plot for IR stability over time, length, PIR UT, GC percentage at PIR 30% cutoff
```{r}
# Define custom colors for timepoints
myColors <- colorRampPalette(c("darkred", "darkgray"))(4)
myStabColors <- colorRampPalette(c("darkred", "darkgray"))(4)

# Prepare long-format data for violin plots
combined_violin_long <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  dplyr::select(GENE, COORD, EVENT, stability,
                Nuc_UT_mean, Nuc_30_mean, Nuc_2h_mean, Nuc_4h_mean) %>%
  pivot_longer(
    cols = starts_with("Nuc_"),
    names_to = "TimePoint",
    values_to = "NuclearMean") %>%
  mutate(
    TimePoint = recode(TimePoint,
                       "Nuc_UT_mean" = "UT",
                       "Nuc_30_mean" = "30min",
                       "Nuc_2h_mean" = "2h",
                       "Nuc_4h_mean" = "4h"),
    TimePoint = factor(TimePoint, levels = c("UT", "30min", "2h", "4h")),
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

# Calculate a shared y-axis range
ymax <- max(combined_violin_long$NuclearMean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10  # Add extra space above the top tick

# Calculate means per TimePoint and stability group
mean_points <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(mean_val = mean(NuclearMean, na.rm = TRUE)) %>%
  ungroup()

# Generate the violin plot with shared scale and extra headroom
p <- ggplot(combined_violin_long, aes(x = TimePoint, y = NuclearMean, fill = TimePoint)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) + #if you want the width to represent actual data density, then geom_violin(trim = FALSE, alpha = 0.7)
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  facet_wrap(~ stability, ncol = 2, scales = "fixed") +
  scale_y_continuous(breaks = ybreaks, limits = c(min(ybreaks), upper_limit)) +
  scale_fill_manual(values = myColors) +
  labs(title = "Violin Plots of Nuclear IR Means by Time Point",
       x = "Time Point", y = "Nuclear Mean PIR") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    strip.text = element_text(face = "bold", size = 13)
  )

ggsave(
  filename = "Figure_2C_IR_stability_timecourse.pdf",
  plot = p,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)
# Count unique EVENTs per stability class
event_counts_ut <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(
    stability = factor(stability,
                       levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  ) %>%
  count(stability, name = "n_EVENT")

event_counts_ut

# Barplot
p_event_bar <- ggplot(event_counts_ut, aes(x = stability, y = n_EVENT, fill = stability)) +
  geom_col(width = 0.7, alpha = 0.9) +
  geom_text(aes(label = n_EVENT), vjust = -0.4, size = 4) +
  scale_fill_manual(values = myStabColors) +
  labs(
    title = "Number of introns per stability class",
    x = "Stability Group",
    y = "Number of introns (EVENTs)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black")
  ) +
  expand_limits(y = max(event_counts_ut$n_EVENT) * 1.1)

ggsave(
  filename = "Suppl_2A_IR_stability_barplot.pdf",
  plot = p_event_bar,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

# Calculate mean NuclearMean PIR per time point and stability group
pir_summary <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(
    mean_PIR = mean(NuclearMean, na.rm = TRUE),
    median_PIR = median(NuclearMean, na.rm = TRUE),
    sd_PIR = sd(NuclearMean, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(stability, TimePoint)

# View the table
print(pir_summary)

write_csv(pir_summary, "Figure_2C_pir_summary.csv")

### NUCLEAR PIR levels
# Calculate y-axis breaks and upper limit
ymax <- max(combined_dataTPMno100$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

p_nuc_ut <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = Nuc_UT_mean, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Nuclear IR (UT)", x = "Stability Group", y = "Nuc_UT_mean") +
  scale_y_continuous(breaks = ybreaks, limits = c(0, upper_limit)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )

ggsave(
  filename = "Fig_2_p_nuc_ut.pdf",
  plot = p_nuc_ut,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

## LENGTH
p_length_log <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"),
         LENGTH > 0) %>%  # Make sure there are no 0-length values
  mutate(
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  ) %>%
  ggplot(aes(x = stability, y = LENGTH, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_log10(limits = c(10, NA), breaks = c(10, 100, 1000, 10000, 100000)) +
  labs(title = "Intron Length (log10 scale)", x = "Stability Group", y = "Length (log10 nt)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  filename = "Fig_2_p_length_log.pdf",
  plot = p_length_log,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

#### GC %
ymax_gc <- max(combined_dataTPMno100$GC_Content, na.rm = TRUE)
ybreaks_gc <- pretty(c(0, ymax_gc))
upper_limit_gc <- max(ybreaks_gc, 100)  # Ensure it reaches at least 100

ymax <- max(combined_dataTPMno100$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10
# Plot
p_gc <- combined_dataTPMno100 %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = GC_Content, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(breaks = ybreaks_gc, limits = c(0, upper_limit)) +
  labs(title = "GC Content", x = "Stability Group", y = "GC (%)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  filename = "Fig_2_p_gc.pdf",
  plot = p_gc,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)
```

# the same plot for 50% cutoff
```{r}
### 1. Prepare long-format data for violin plots
combined_violin_long <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  dplyr::select(GENE, stability, Nuc_UT_mean, Nuc_30_mean, Nuc_2h_mean, Nuc_4h_mean) %>%
  pivot_longer(
    cols = starts_with("Nuc_"),
    names_to = "TimePoint",
    values_to = "NuclearMean"
  ) %>%
  mutate(
    TimePoint = recode(TimePoint,
                       "Nuc_UT_mean" = "UT",
                       "Nuc_30_mean" = "30min",
                       "Nuc_2h_mean" = "2h",
                       "Nuc_4h_mean" = "4h"),
    TimePoint = factor(TimePoint, levels = c("UT", "30min", "2h", "4h")),
    stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  )

# Calculate shared y-axis
ymax <- max(combined_violin_long$NuclearMean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

# Calculate means for dot overlay
mean_points <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(mean_val = mean(NuclearMean, na.rm = TRUE)) %>%
  ungroup()

# Plot nuclear PIR violin plot
p_50 <- ggplot(combined_violin_long, aes(x = TimePoint, y = NuclearMean, fill = TimePoint)) +
  geom_violin(trim = FALSE, scale = "width", alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  facet_wrap(~ stability, ncol = 2, scales = "fixed") +
  scale_y_continuous(breaks = ybreaks, limits = c(min(ybreaks), upper_limit)) +
  scale_fill_manual(values = myColors) +
  labs(title = "Violin Plots of Nuclear IR Means by Time Point",
       x = "Time Point", y = "Nuclear Mean PIR") +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    strip.text = element_text(face = "bold", size = 13)
  )

# Save plot
ggsave(
  filename = "Figure_2_p_50.pdf",
  plot = p_50,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

# Summary table
pir_summary_50cut <- combined_violin_long %>%
  group_by(stability, TimePoint) %>%
  summarise(
    mean_PIR = mean(NuclearMean, na.rm = TRUE),
    median_PIR = median(NuclearMean, na.rm = TRUE),
    sd_PIR = sd(NuclearMean, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(stability, TimePoint)

print(pir_summary_50cut)
write_csv(pir_summary_50cut, "Figure_2_pir_summary_50cut.csv")

### 2. Basal Nuclear PIR Plot
ymax <- max(combined_dataTPMno100_50cut$Nuc_UT_mean, na.rm = TRUE)
ybreaks <- pretty(c(0, ymax))
upper_limit <- max(ybreaks) + 10

p_nuc_ut_50 <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = Nuc_UT_mean, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  labs(title = "Nuclear IR (UT)", x = "Stability Group", y = "Nuc_UT_mean") +
  scale_y_continuous(breaks = ybreaks, limits = c(0, upper_limit)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "none"
  )

ggsave(
  filename = "Figure_2_p_nuc_ut_50.pdf",
  plot = p_nuc_ut_50,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

### 3. Intron Length (log10)
p_length_log_50 <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"),
         LENGTH > 0) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = LENGTH, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_log10(limits = c(10, NA), breaks = c(10, 100, 1000, 10000, 100000)) +
  labs(title = "Intron Length (log10 scale)", x = "Stability Group", y = "Length (log10 nt)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  filename = "Figure_2_p_length_log_50.pdf",
  plot = p_length_log_50,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

### 4. GC Content
ymax_gc <- max(combined_dataTPMno100_50cut$GC_Content, na.rm = TRUE)
ybreaks_gc <- pretty(c(0, ymax_gc))
upper_limit_gc <- max(ybreaks_gc, 100)

p_gc_50 <- combined_dataTPMno100_50cut %>%
  filter(stability %in% c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")) %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))) %>%
  ggplot(aes(x = stability, y = GC_Content, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.8) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_y_continuous(breaks = ybreaks_gc, limits = c(0, upper_limit_gc)) +
  labs(title = "GC Content", x = "Stability Group", y = "GC (%)") +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

ggsave(
  filename = "Figure_2_p_gc_50.pdf",
  plot = p_gc_50,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)
```

#############
# Classify genes into stable and unstable groups, 30% PIR cutoff
#############
```{r}
# Check overlapping introns between groups
sets_list <- list(
  "Stable 4h" = unique(stableR4h$COORD),
  "Unstable 4h" = unique(unstableR4h$COORD),
  "Unstable 30min" = unique(unstableR30min$COORD),
  "Unstable 2h" = unique(unstableR2h$COORD),
  "No IR" = unique(noIR$COORD))

sets_list <- list(
  "Stable 4h" = unique(stableR4h$GENE),
  "Unstable 4h" = unique(unstableR4h$GENE),
  "Unstable 30min" = unique(unstableR30min$GENE),
  "Unstable 2h" = unique(unstableR2h$GENE))

exclusive_by_set <- lapply(names(sets_list), function(set_name) {
  current_set <- sets_list[[set_name]]
  # Combine all the genes in the other sets into one vector:
  other_sets <- unlist(sets_list[names(sets_list) != set_name])
  setdiff(current_set, other_sets)})
names(exclusive_by_set) <- names(sets_list)

non_overlapping_genes <- unlist(exclusive_by_set)
non_overlapping_genes

exclusive_unstable30min <- setdiff(
  sets_list[["Unstable 30min"]],
  union(union(sets_list[["Stable 4h"]], sets_list[["Unstable 4h"]]), sets_list[["Unstable 2h"]]))

exclusive_unstable2h <- setdiff(
  sets_list[["Unstable 2h"]],
  union(sets_list[["Stable 4h"]], sets_list[["Unstable 4h"]]))

exclusive_unstable4h <- setdiff(unstableR4h$GENE, stableR4h$GENE)

sets_list_updated <- list(
  "Stable 4h"      = unique(stableR4h$GENE),
  "Unstable 4h"    = unique(exclusive_unstable4h),
  "Unstable 30min" = unique(exclusive_unstable30min),
  "Unstable 2h"    = unique(exclusive_unstable2h))

initial_genes <- unique(c(
  stableR4h$GENE,
  unstableR4h$GENE,
  unstableR30min$GENE,
  unstableR2h$GENE))

filtered_genes <- unique(c(
  stableR4h$GENE,
  exclusive_unstable4h,              # Already removed genes overlapping stableR4h
  exclusive_unstable30min,
  exclusive_unstable2h))

# Print out the total counts for comparison
cat("Total unique genes before filtering:", length(initial_genes), "\n")
cat("Total unique genes after filtering: ", length(filtered_genes), "\n")

# You can also check that the sets are identical (if the overall union remains unchanged)
if (setequal(initial_genes, filtered_genes)) {
  cat("The union of genes is identical before and after filtering.\n")
} else {
  cat("The union of genes differs between the initial and filtered sets.\n")}

# Create labeled data frame from stableR4h (which is a data frame)
df_stable4h <- stableR4h %>%
  dplyr::select(GENE) %>%
  mutate(GENE_stability = "stable_4h")

# For the others, turn them into data frames directly
df_unstable4h <- data.frame(GENE = exclusive_unstable4h,
                            GENE_stability = "unstable_4h")

df_unstable30min <- data.frame(GENE = exclusive_unstable30min,
                               GENE_stability = "unstable_30min")

df_unstable2h <- data.frame(GENE = exclusive_unstable2h,
                            GENE_stability = "unstable_2h")

# Combine all into one data frame
gene_stability_df <- bind_rows(
  df_stable4h,
  df_unstable4h,
  df_unstable30min,
  df_unstable2h)
gene_stability_df_unique <- gene_stability_df %>%
  distinct(GENE, .keep_all = TRUE)  # Keep only one row per gene
```

# Classify genes into stable and unstable groups, 50% PIR cutoff
```{r}
sets_list_50cut_coord <- list(
  "Stable 4h"       = unique(stableR4h_50cut$COORD),
  "Unstable 4h"     = unique(unstableR4h_50cut$COORD),
  "Unstable 30min"  = unique(unstableR30min_50cut$COORD),
  "Unstable 2h"     = unique(unstableR2h_50cut$COORD),
  "No IR"           = unique(noIR_50cut$COORD)
)

sets_list_50cut_gene <- list(
  "Stable 4h"       = unique(stableR4h_50cut$GENE),
  "Unstable 4h"     = unique(unstableR4h_50cut$GENE),
  "Unstable 30min"  = unique(unstableR30min_50cut$GENE),
  "Unstable 2h"     = unique(unstableR2h_50cut$GENE)
  # "No IR" can be included if desired
)

exclusive_by_set_50cut <- lapply(names(sets_list_50cut_gene), function(set_name) {
  current_set <- sets_list_50cut_gene[[set_name]]
  other_sets <- unlist(sets_list_50cut_gene[names(sets_list_50cut_gene) != set_name])
  setdiff(current_set, other_sets)
})
names(exclusive_by_set_50cut) <- names(sets_list_50cut_gene)

non_overlapping_genes_50cut <- unlist(exclusive_by_set_50cut)

exclusive_unstable30min_50cut <- setdiff(
  sets_list_50cut_gene[["Unstable 30min"]],
  union(union(sets_list_50cut_gene[["Stable 4h"]], sets_list_50cut_gene[["Unstable 4h"]]),
        sets_list_50cut_gene[["Unstable 2h"]])
)

exclusive_unstable2h_50cut <- setdiff(
  sets_list_50cut_gene[["Unstable 2h"]],
  union(sets_list_50cut_gene[["Stable 4h"]], sets_list_50cut_gene[["Unstable 4h"]])
)

exclusive_unstable4h_50cut <- setdiff(
  unstableR4h_50cut$GENE,
  stableR4h_50cut$GENE
)

sets_list_updated_50cut <- list(
  "Stable 4h"      = unique(stableR4h_50cut$GENE),
  "Unstable 4h"    = unique(exclusive_unstable4h_50cut),
  "Unstable 30min" = unique(exclusive_unstable30min_50cut),
  "Unstable 2h"    = unique(exclusive_unstable2h_50cut)
)

initial_genes_50cut <- unique(c(
  stableR4h_50cut$GENE,
  unstableR4h_50cut$GENE,
  unstableR30min_50cut$GENE,
  unstableR2h_50cut$GENE
))

filtered_genes_50cut <- unique(c(
  stableR4h_50cut$GENE,
  exclusive_unstable4h_50cut,
  exclusive_unstable30min_50cut,
  exclusive_unstable2h_50cut
))

cat("Total unique genes before filtering (50%):", length(initial_genes_50cut), "\n")
cat("Total unique genes after filtering (50%): ", length(filtered_genes_50cut), "\n")

if (setequal(initial_genes_50cut, filtered_genes_50cut)) {
  cat("The union of genes is identical before and after filtering.\n")
} else {
  cat("The union of genes differs between the initial and filtered sets.\n")
}

df_stable4h_50cut <- stableR4h_50cut %>%
  dplyr::select(GENE) %>%
  mutate(GENE_stability = "stable_4h")

df_unstable4h_50cut <- data.frame(GENE = exclusive_unstable4h_50cut,
                                  GENE_stability = "unstable_4h")

df_unstable30min_50cut <- data.frame(GENE = exclusive_unstable30min_50cut,
                                     GENE_stability = "unstable_30min")

df_unstable2h_50cut <- data.frame(GENE = exclusive_unstable2h_50cut,
                                  GENE_stability = "unstable_2h")

gene_stability_df_50cut <- bind_rows(
  df_stable4h_50cut,
  df_unstable4h_50cut,
  df_unstable30min_50cut,
  df_unstable2h_50cut
)

gene_stability_df_unique_50cut <- gene_stability_df_50cut %>%
  distinct(GENE, .keep_all = TRUE)
```

### Line plots of RNA stability and nuclear enrichment at the 30% PIR cutoff ###
```{r}
# For RNA stability analyses in the nucleus and cytoplasm, we applied expression
# thresholds to remove low-abundance transcripts. To assess cytoplasmic RNA
# stability, we required Cyt and Nuc UT_TPM_mean >= 7. This threshold was based on RAD52
# smRNA FISH measurements, as RAD52 showed the lowest cytoplasmic abundance among
# the IR mRNAs analyzed, corresponding to a mean of approximately one molecule
# per cytoplasm per cell.

combined_dataTPMno100 <- combined_data %>%
 filter(Nuc_UT_TPM_mean >= 7 & Cyt_UT_TPM_mean >= 7) %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | #keep is.na because otherwise they get moved
   (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

combined_exclusive <- combined_dataTPMno100 %>%
  filter(GENE %in% gene_stability_df_unique$GENE) %>%
  left_join(gene_stability_df_unique, by = "GENE") %>%
  filter(stability != "no IR")
sum(duplicated(combined_exclusive$EVENT))

combined_exclusive_unique <- combined_exclusive %>%
  distinct(GENE, .keep_all = TRUE)
sum(duplicated(combined_exclusive_unique$GENE))

stable4h_genes_only <- combined_exclusive_unique %>%
  filter(GENE_stability == "stable_4h")

subset_combined_exclusive <- combined_exclusive %>%
  dplyr::select(
    GENE, COORD, LENGTH,
    Cyt_A_UT_1.x, Cyt_B_UT_1.x, Cyt_A_30_1.x, Cyt_B_30_1.x, 
    Cyt_A_2h_1.x, Cyt_B_2h_1.x, Cyt_A_4h_1.x, Cyt_B_4h_1.x,
    Nuc_A_UT_1.x, Nuc_B_UT_1.x, Nuc_A_30_1.x, Nuc_B_30_1.x,
    Nuc_A_2h_1.x, Nuc_B_2h_1.x, Nuc_A_4h_1.x, Nuc_B_4h_1.x,
    Cyt_UT_mean, Cyt_30_mean, Cyt_2h_mean, Cyt_4h_mean,
    Nuc_UT_mean, Nuc_UT_TPM_mean, Nuc_30_TPM_mean, Nuc_2h_TPM_mean, Nuc_4h_TPM_mean,
    Cyt_UT_TPM_mean, Cyt_30_TPM_mean, Cyt_2h_TPM_mean, Cyt_4h_TPM_mean,
    Nuc_UT_IR_stability, Nuc_30_IR_stability, Nuc_2h_IR_stability, Nuc_4h_IR_stability,
    Cyt_UT_IR_stability, Cyt_30_IR_stability, Cyt_2h_IR_stability, Cyt_4h_IR_stability,
    stability, GENE_stability)

long_exclusive_data <- combined_exclusive_unique %>%
  pivot_longer(
    cols = c("Nuc_UT_TPM_stability", "Nuc_30_TPM_stability", 
             "Nuc_2h_TPM_stability", "Nuc_4h_TPM_stability",
             "Cyt_UT_TPM_stability", "Cyt_30_TPM_stability", 
             "Cyt_2h_TPM_stability", "Cyt_4h_TPM_stability"),
    names_to = c("Compartment", "Time"),
    names_pattern = "^(Nuc|Cyt)_([^_]+)_TPM_stability$",
    values_to = "TPM_stability") %>%
  mutate(Time = factor(Time, levels = c("UT", "30", "2h", "4h")))

long_exclusive_data <- long_exclusive_data %>%
  filter(stability != "no IR") %>%  # Remove rows with "no IR"
  mutate(stability = factor(as.character(stability),
                            levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

myStabColors <- colorRampPalette(c("darkgray", "darkred"))(4)

l_30 <- ggplot(long_exclusive_data, aes(x = Time, y = TPM_stability, color = GENE_stability, group = GENE_stability)) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = GENE_stability), alpha = 0.2, color = NA) +  # Add ribbon first
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ Compartment, scales = "free_y") +
  labs(title = "RNA Stability Over Time (Exclusive Gene Groups)",
       x = "Time Point",
       y = "Mean TPM Stability (Relative to UT)") +
  coord_cartesian(ylim = c(0.0, 1.05)) +
  scale_color_manual(
    breaks = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h"),
    values = myStabColors) +
  scale_fill_manual(  # Add this to match ribbon fill colors
    breaks = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h"),
    values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

ggsave(
  filename = "Figure_2_l_30.pdf",
  plot = l_30,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

stability_summary <- long_exclusive_data %>%
  group_by(GENE_stability, Time, Compartment) %>%
  summarise(
    mean_TPM_stability   = mean(TPM_stability, na.rm = TRUE),
    median_TPM_stability = median(TPM_stability, na.rm = TRUE),
    sd_TPM_stability     = sd(TPM_stability, na.rm = TRUE),
    n                    = sum(!is.na(TPM_stability)),
    .groups = "drop"
  )

# ------------------------------------------------------------------------------
# 2. Safe Welch t-test helper
# ------------------------------------------------------------------------------

safe_welch_test <- function(x, y) {
  x <- x[!is.na(x)]
  y <- y[!is.na(y)]
  
  n1 <- length(x)
  n2 <- length(y)
  
  sd1 <- if (n1 > 1) sd(x) else NA_real_
  sd2 <- if (n2 > 1) sd(y) else NA_real_
  
  # not enough observations
  if (n1 < 2 || n2 < 2) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: fewer than 2 observations in at least one group"
    ))
  }
  
  # constant groups
  if (!is.na(sd1) && !is.na(sd2) && sd1 == 0 && sd2 == 0) {
    if (mean(x) == mean(y)) {
      return(tibble(
        statistic = NA_real_,
        p_value   = 1,
        method    = "Both groups constant and identical"
      ))
    } else {
      return(tibble(
        statistic = NA_real_,
        p_value   = NA_real_,
        method    = "Welch's t-test not computed: both groups constant but with different means"
      ))
    }
  }
  
  if (!is.na(sd1) && sd1 == 0) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: group1 is constant"
    ))
  }
  
  if (!is.na(sd2) && sd2 == 0) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: group2 is constant"
    ))
  }
  
  # try Welch t-test
  out <- tryCatch(
    {
      tt <- t.test(x, y, var.equal = FALSE, alternative = "two.sided")
      tibble(
        statistic = unname(tt$statistic),
        p_value   = tt$p.value,
        method    = tt$method
      )
    },
    error = function(e) {
      tibble(
        statistic = NA_real_,
        p_value   = NA_real_,
        method    = paste("Welch's t-test failed:", e$message)
      )
    }
  )
  
  out
}

# ---  Calculate the log2 Fold Change for Nuclear Enrichment ---
combined_dataTPMno100 <- combined_data %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) |
   (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

combined_exclusive <- combined_dataTPMno100 %>%
  filter(GENE %in% gene_stability_df_unique$GENE) %>%
  left_join(gene_stability_df_unique, by = "GENE") %>%
  filter(stability != "no IR")
sum(duplicated(combined_exclusive$GENE))

combined_exclusive_unique <- combined_exclusive %>%
  distinct(GENE, .keep_all = TRUE)
sum(duplicated(combined_exclusive_unique$GENE))

combined_exclusive_filtered <- combined_exclusive_unique %>%
  #filter(stability != "no IR") %>%
  mutate(
    nuc_enrichment_log2FC = log2(Nuc_UT_TPM_mean / Cyt_UT_TPM_mean),
    stability = factor(as.character(stability),
                       levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h"))
  )

p3 <- ggplot(combined_exclusive_filtered, aes(x = stability, y = nuc_enrichment_log2FC, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.7) +                
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  labs(
    title = "Violin Plot of Nuclear Enrichment log2FC",
    x = "Stability Group",
    y = "log2FC (Nuc_UT_TPM_mean / Cyt_UT_TPM_mean)"
  ) +
  scale_y_continuous(limits = c(-2.5, NA), expand = c(0, 0.05)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

ggsave(
  filename = "Figure_2_p3_30.pdf",
  plot = p3,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)
```

### Line plots of RNA stability and nuclear enrichment at the 50% PIR cutoff ###
```{r}
combined_dataTPMno100_50cut <- combined_data_50cut %>%
  filter(Nuc_UT_TPM_mean >= 7 & Cyt_UT_TPM_mean >= 7) %>%
  filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) |
         (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

combined_exclusive_50cut <- combined_dataTPMno100_50cut %>%
  filter(GENE %in% gene_stability_df_unique_50cut$GENE) %>%
  left_join(gene_stability_df_unique_50cut, by = "GENE") %>%
  filter(stability != "no IR")

combined_exclusive_unique_50cut <- combined_exclusive_50cut %>%
  distinct(GENE, .keep_all = TRUE)

subset_combined_exclusive_50cut <- combined_exclusive_50cut %>%
  dplyr::select(
    GENE, COORD, LENGTH,
    Cyt_A_UT_1.x, Cyt_B_UT_1.x, Cyt_A_30_1.x, Cyt_B_30_1.x, 
    Cyt_A_2h_1.x, Cyt_B_2h_1.x, Cyt_A_4h_1.x, Cyt_B_4h_1.x,
    Nuc_A_UT_1.x, Nuc_B_UT_1.x, Nuc_A_30_1.x, Nuc_B_30_1.x,
    Nuc_A_2h_1.x, Nuc_B_2h_1.x, Nuc_A_4h_1.x, Nuc_B_4h_1.x,
    Cyt_UT_mean, Cyt_30_mean, Cyt_2h_mean, Cyt_4h_mean,
    Nuc_UT_mean, Nuc_UT_TPM_mean, Nuc_30_TPM_mean, Nuc_2h_TPM_mean, Nuc_4h_TPM_mean,
    Cyt_UT_TPM_mean, Cyt_30_TPM_mean, Cyt_2h_TPM_mean, Cyt_4h_TPM_mean,
    Nuc_UT_IR_stability, Nuc_30_IR_stability, Nuc_2h_IR_stability, Nuc_4h_IR_stability,
    Cyt_UT_IR_stability, Cyt_30_IR_stability, Cyt_2h_IR_stability, Cyt_4h_IR_stability,
    stability, GENE_stability)

long_exclusive_data_50cut <- combined_exclusive_unique_50cut %>%
  pivot_longer(
    cols = c("Nuc_UT_TPM_stability", "Nuc_30_TPM_stability", 
             "Nuc_2h_TPM_stability", "Nuc_4h_TPM_stability",
             "Cyt_UT_TPM_stability", "Cyt_30_TPM_stability", 
             "Cyt_2h_TPM_stability", "Cyt_4h_TPM_stability"),
    names_to = c("Compartment", "Time"),
    names_pattern = "^(Nuc|Cyt)_([^_]+)_TPM_stability$",
    values_to = "TPM_stability") %>%
  mutate(Time = factor(Time, levels = c("UT", "30", "2h", "4h")),
         stability = factor(as.character(stability),
                            levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

long_exclusive_data_50cut$GENE_stability <- factor(
  long_exclusive_data_50cut$GENE_stability,
  levels = c("unstable_30min", "unstable_2h", "unstable_4h", "stable_4h")
)

long_exclusive_data_50cut %>%
  group_by(stability) %>%
  summarize(n_genes = n_distinct(GENE)) %>%
  print()

myStabColors <- colorRampPalette(c("darkgray", "darkred"))(4)
l_50cut <- ggplot(long_exclusive_data_50cut, aes(x = Time, y = TPM_stability, color = GENE_stability, group = GENE_stability)) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = GENE_stability), alpha = 0.2, color = NA) +
  stat_summary(fun = mean, geom = "line", size = 1.2) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  facet_wrap(~ Compartment, scales = "free_y") +
  labs(title = "RNA Stability Over Time (Exclusive Gene Groups, 50% cutoff)",
       x = "Time Point", y = "Mean TPM Stability (Relative to UT)") +
  coord_cartesian(ylim = c(0.0, 1.05)) +
  scale_color_manual(values = myStabColors) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.x = element_line(color = "black"),
    axis.line.y = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16)
  )

# Display the updated plot
ggsave(
  filename = "Figure_2_l_50cut.pdf",
  plot = l_50cut,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)

stability_summary_50 <- long_exclusive_data_50cut %>%
  group_by(GENE_stability, Time, Compartment) %>%
  summarise(
    mean_TPM_stability   = mean(TPM_stability, na.rm = TRUE),
    median_TPM_stability = median(TPM_stability, na.rm = TRUE),
    sd_TPM_stability     = sd(TPM_stability, na.rm = TRUE),
    n                    = sum(!is.na(TPM_stability)),
    .groups = "drop"
  )

# ------------------------------------------------------------------------------
# 2. Safe Welch t-test helper
# ------------------------------------------------------------------------------

safe_welch_test <- function(x, y) {
  x <- x[!is.na(x)]
  y <- y[!is.na(y)]

  n1 <- length(x)
  n2 <- length(y)

  sd1 <- if (n1 > 1) sd(x) else NA_real_
  sd2 <- if (n2 > 1) sd(y) else NA_real_

  if (n1 < 2 || n2 < 2) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: fewer than 2 observations in at least one group"
    ))
  }

  if (!is.na(sd1) && !is.na(sd2) && sd1 == 0 && sd2 == 0) {
    if (mean(x) == mean(y)) {
      return(tibble(
        statistic = NA_real_,
        p_value   = 1,
        method    = "Both groups constant and identical"
      ))
    } else {
      return(tibble(
        statistic = NA_real_,
        p_value   = NA_real_,
        method    = "Welch's t-test not computed: both groups constant but with different means"
      ))
    }
  }

  if (!is.na(sd1) && sd1 == 0) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: group1 is constant"
    ))
  }

  if (!is.na(sd2) && sd2 == 0) {
    return(tibble(
      statistic = NA_real_,
      p_value   = NA_real_,
      method    = "Welch's t-test not computed: group2 is constant"
    ))
  }

  out <- tryCatch(
    {
      tt <- t.test(x, y, var.equal = FALSE, alternative = "two.sided")
      tibble(
        statistic = unname(tt$statistic),
        p_value   = tt$p.value,
        method    = tt$method
      )
    },
    error = function(e) {
      tibble(
        statistic = NA_real_,
        p_value   = NA_real_,
        method    = paste("Welch's t-test failed:", e$message)
      )
    }
  )

  out
}

# ------------------------------------------------------------------------------------------------------
# Violin plot of log2 nuclear enrichment
combined_dataTPMno100_50cut <- combined_data_50cut %>%
   filter(is.na(Cyt_A_UT_1.x) | is.na(Cyt_B_UT_1.x) | 
         (Cyt_A_UT_1.x < 99 & Cyt_B_UT_1.x < 99))

combined_exclusive_50cut <- combined_dataTPMno100_50cut %>%
  filter(GENE %in% gene_stability_df_unique_50cut$GENE) %>%
  left_join(gene_stability_df_unique_50cut, by = "GENE") %>%
  filter(stability != "no IR")

combined_exclusive_unique_50cut <- combined_exclusive_50cut %>%
  distinct(GENE, .keep_all = TRUE)

combined_exclusive_filtered_50cut <- combined_exclusive_unique_50cut %>%
  filter(stability != "no IR") %>%
  mutate(nuc_enrichment_log2FC = log2(Nuc_UT_TPM_mean / Cyt_UT_TPM_mean))

combined_exclusive_filtered_50cut <- combined_exclusive_filtered_50cut %>%
  mutate(stability = factor(stability, levels = c("unstable30min", "unstable2h", "unstable 4h", "stable 4h")))

p3_50cut <- ggplot(combined_exclusive_filtered_50cut, aes(x = stability, y = nuc_enrichment_log2FC, fill = stability)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  labs(title = "Violin Plot of Nuclear Enrichment log2FC (50% cutoff)",
       x = "Stability Group", y = "log2FC (Nuc_UT_TPM / Cyt_UT_TPM)") +
  scale_y_continuous(limits = c(-2.5, NA), expand = c(0, 0.05)) +
  scale_fill_manual(values = myStabColors) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 16))

ggsave(
  filename = "Figure_2_p3_50cut.pdf",
  plot = p3_50cut,
  device = "pdf",
  width = 10,
  height = 8,
  units = "in"
)
```